Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-18 paper/paper.qmd
Clarify connectivity particle release
Minor fix to formatting
Tweak wording to address concerns
Also adds a reference

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of climatic and ecological conditions. Exploratory modelling
approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] to address
limitations of calibrated ecosystem models that use a singular set of parameter values
[@lubianabotelho2025]. Exploratory scenario modelling is a core methodology underpinning the
development of robust and adaptive management strategies [@haasnoot2013].

All data preparation, modelling and analyses were performed using the Julia programming
language [@bezanson2017], with visualisations created using the Makie plotting ecosystem
[@danisch2021]. Code used to process DHW data and perform analyses involved in this
investigation can be found at Grier and Iwanaga [-@grier2025].

@@

For each spawning night, particles were released from random locations within each reef
polygon as neutrally buoyant virtual larvae every 27 seconds from 8:00 pm to 11:00 pm at
2.25 m below the surface. In each of the reef polygons, 401 particles were released
resulting in a total of 1,526,206 particles every night of spawning. The model was run for
30 days with a 3-minute timestep, and the locations of the released particles were updated
every 15 minutes until the end of the simulation. Larvae were assumed to become competent to
settle from 4 to 28 days post release with a daily mortality rate of 0.4 [@ani2024], which
was implemented in 15-minute timestep as probabilistic mortality, corresponding to 0.0053
mortality rate per timestep. To incorporate the stochastic sensory behaviour of larvae
during settlement, a 0.01 probability of settlement is assumed when larvae are over a reef
during the competency period per 15-minute timestep. The information obtained on larvae
settlement was used to compute a connectivity matrix among reef polygons for each spawning
night (see Ani et al. [-@ani2024] for more details). A total of 24 connectivity matrices
were generated for the spawning nights that were considered.

@@

Two systems were employed to explore scenarios: ADRIA (the Adaptive Dynamic Reef
Intervention Algorithms platform) and CoralBlox, referred to as the coupled framework
ADRIA-CoralBlox. ADRIA is a multi-criteria decision support platform designed for the
assessment and analysis of coral reef intervention strategies under uncertainty
[@iwanaga2023]. The ADRIA platform provides functionality for scenario generation and
analysis as well as providing technical infrastructure to couple environmental and ecosystem
representations. CoralBlox is a computationally efficient coral population model. It
simulates functional/morphological groups of corals over their life cycles, competing for
available space over time [@ribeirodealmeida2024].

>>>>

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of climatic and ecological conditions. Exploratory modelling
approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013]. Exploratory
scenario modelling is a core methodology underpinning the development of robust and adaptive
management strategies [@haasnoot2013].

Recent critiques have raised concerns around the use of calibrated ecological models - which
often have poor predictive capacity - for decision support [@lubianabotelho2025]. Here, the
focus is on exploring the climatic and physical context that pertain to management decisions
to understand the range of conditions under which conclusions hold true, rather than
providing precise forecasts of future state. These environmental factors often set the
context which influences ecological outcomes and have more manageable uncertainties than the
precise ecological parameterisations [@reichert2005].

All data preparation, modelling and analyses were performed using the Julia programming
language [@bezanson2017], with visualisations created using the Makie plotting ecosystem
[@danisch2021]. Code used to process DHW data, perform analyses, and create figures for this
investigation are openly published and can be found in Grier and Iwanaga [-@grier2025].

@@

For each spawning night, particles were released from random locations within each reef
polygon as neutrally buoyant virtual larvae every 27 seconds from 8:00 pm to 11:00 pm at
2.25 m below the surface. In each of the reef polygons, 401 particles were released
resulting in a total of 1,526,206 particles every night of spawning across the GBR. The
model was run for 30 days with a 3-minute timestep, and the locations of the released
particles were updated every 15 minutes until the end of the simulation. Larvae were assumed
to become competent to settle from 4 to 28 days post release with a daily mortality rate of
0.4 [@ani2024], which was implemented in 15-minute timestep as probabilistic mortality,
corresponding to 0.0053 mortality rate per timestep. To incorporate the stochastic sensory
behaviour of larvae during settlement, a 0.01 probability of settlement is assumed when
larvae are over a reef during the competency period per 15-minute timestep. The information
obtained on larvae settlement was used to compute a connectivity matrix among reef polygons
for each spawning night (see Ani et al. [-@ani2024] for more details). A total of 24
connectivity matrices were generated for the spawning nights that were considered.

@@

Two modelling systems were employed to explore scenarios: ADRIA (the Adaptive Dynamic Reef
Intervention Algorithms platform) and CoralBlox, referred to as the coupled framework
ADRIA-CoralBlox. ADRIA is a multi-criteria decision support platform designed for the
assessment and analysis of coral reef intervention strategies under uncertainty
[@iwanaga2023]. The ADRIA platform provides functionality for scenario generation and
analysis as well as providing technical infrastructure to couple environmental and ecosystem
representations. CoralBlox is a computationally efficient coral population model. It
simulates functional/morphological groups of corals over their life cycles, competing for
available space over time [@ribeirodealmeida2024].
